```{r mds-tables}
# mds' own report, knitted by the frame's report.Rmd with the functions of
# analysis/lib/rmd/report.R. It reads the tables this module published and nothing else.
#
# Pools keep mds.tsv's order.
coords <- report_read("mds.tsv")
pairs <- report_read("distance.tsv")
eigen <- report_read("eigenvalues.tsv")
num <- function(x) suppressWarnings(as.numeric(x))

pools <- coords$pool
axes <- grep("^dim[0-9]+$", names(coords), value = TRUE)
variables <- setdiff(names(coords), c("pool", "unit", axes))
units <- length(unique(coords$unit))
shared <- num(pairs$sites)
absolute <- num(eigen$share_absolute)
drawn <- num(eigen$plotted) == 1
carried <- if (any(drawn)) {
    paste0("the two axes drawn carry ", report_number(100 * sum(absolute[drawn]), decimals = 1),
           "% of the absolute eigenvalue sum")
} else {
    paste0("the one axis written carries ", report_number(100 * absolute[1], decimals = 1),
           "% of the absolute eigenvalue sum, and one axis draws no plot")
}
# An eigenvalue within rounding of zero counts as zero: n pools have at most n - 1 axes, and the
# last eigenvalue lands either side of zero.
eigenvalues <- num(eigen$eigenvalue)
negative <- !is.na(eigenvalues) & eigenvalues < -1e-10 * max(abs(eigenvalues), na.rm = TRUE)
negative_share <- sum(abs(eigenvalues[negative])) / sum(abs(eigenvalues), na.rm = TRUE)

# One pool-by-pool matrix out of the pair rows, from `column`, with the diagonal left empty.
pool_matrix <- function(column, ...) {
    held <- matrix("", length(pools), length(pools), dimnames = list(pools, pools))
    values <- report_number(pairs[[column]], ...)
    held[cbind(match(pairs$pool_a, pools), match(pairs$pool_b, pools))] <- values
    held[cbind(match(pairs$pool_b, pools), match(pairs$pool_a, pools))] <- values
    held
}

# The pools a matrix holds as columns, at most eight to a table.
column_groups <- report_columns(pools)
ends <- cumsum(lengths(column_groups))
matrix_tables <- function(held, label, caption) {
    for (k in seq_along(column_groups)) {
        group <- column_groups[[k]]
        table <- data.frame(Pool = pools, held[, group, drop = FALSE], check.names = FALSE,
                            stringsAsFactors = FALSE)
        which <- if (length(column_groups) > 1) sprintf(" Pools %d to %d of %d as columns.",
                                                         ends[k] - length(group) + 1, ends[k],
                                                         length(pools)) else ""
        report_table(table, groups = c("", rep(label, length(group))),
                     caption = paste0(caption, which))
    }
}
```

`r length(pools)` pools from `r units` `r if (units == 1) "unit" else "units"`. Each distance is averaged over the sites both of its pools were read at, `r if (min(shared) == max(shared)) paste(report_number(min(shared)), "a pair") else paste("from", report_number(min(shared)), "to", report_number(max(shared)), "a pair")`, and `r carried`.

## The pools on the leading axes

```{r mds-plot, results = "asis"}
for (path in report_files("mds.png")) {
    report_image(path, caption = paste0("The pools on the leading two axes (mds.png, drawn from ",
                                        "mds.tsv and eigenvalues.tsv)."))
}
table <- data.frame(Pool = pools, Unit = coords$unit, check.names = FALSE,
                    stringsAsFactors = FALSE)
for (v in variables) table[[v]] <- coords[[v]]
for (a in axes) table[[sub("^dim", "Axis ", a)]] <- report_number(coords[[a]], digits = 3)
report_table(table, caption = paste0("Each pool's coordinate on every axis the run wrote (mds.tsv). ",
                                     "Each axis is signed so that its largest coordinate is positive."))
```

## How much each axis carries

```{r mds-eigenvalues, results = "asis"}
percent <- function(x) paste0(report_number(100 * num(x), decimals = 1), "%")
table <- data.frame(Axis = eigen$axis,
                    Eigenvalue = report_number(eigen$eigenvalue, digits = 3),
                    "Share" = percent(eigen$share),
                    "Share of |sum|" = percent(eigen$share_absolute),
                    "Cumulative |sum|" = percent(eigen$cumulative_absolute),
                    Drawn = ifelse(drawn, "yes", ""),
                    check.names = FALSE, stringsAsFactors = FALSE)
report_table(table, caption = paste0(
    "Every eigenvalue with its share of the scatter (eigenvalues.tsv, where the shares are ",
    "fractions). Share divides by the signed sum and Share of |sum| by the sum of absolute values, ",
    "which the plot's axis labels use. ",
    if (any(negative)) paste0("Negative eigenvalues carry ",
                              report_number(100 * negative_share, decimals = 1),
                              "% of the absolute sum.")
    else if (any(eigenvalues < 0, na.rm = TRUE))
        paste0("No eigenvalue is negative beyond rounding: ",
               report_number(min(eigenvalues, na.rm = TRUE), digits = 3), " is zero.")
    else "No eigenvalue is negative."))

scree <- data.frame(axis = factor(eigen$axis, levels = eigen$axis), share = 100 * absolute,
                    sign = ifelse(negative, "negative", "positive"), stringsAsFactors = FALSE)
figure <- ggplot2::ggplot(scree, ggplot2::aes(x = axis, y = share, fill = sign)) +
    ggplot2::geom_col(width = 0.7) +
    ggplot2::scale_fill_manual(values = c(positive = "#4d4d4d", negative = "#d6604d")) +
    ggplot2::labs(x = "Axis", y = "% of the absolute sum", fill = NULL) +
    ggplot2::theme_bw(base_size = 9) +
    ggplot2::theme(legend.position = "bottom")
report_plot(figure, "eigenvalues", width = 6, height = 2.8,
            caption = "How much each axis carries, from eigenvalues.tsv.")
```

## Distances between pools

```{r mds-distances, results = "asis"}
matrix_tables(pool_matrix("distance", digits = 3), "Corrected distance",
              paste0("Nei's minimum distance between every two pools, corrected for sampling ",
                     "(distance.tsv). A negative entry reads as indistinguishable at this depth; ",
                     "distances are not floored at zero."))

grid <- data.frame(a = factor(c(pairs$pool_a, pairs$pool_b), levels = pools),
                   b = factor(c(pairs$pool_b, pairs$pool_a), levels = rev(pools)),
                   value = num(c(pairs$distance, pairs$distance)))
figure <- ggplot2::ggplot(grid, ggplot2::aes(x = a, y = b, fill = value)) +
    ggplot2::geom_tile(color = "white") +
    ggplot2::scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", midpoint = 0,
                                  name = "Corrected\ndistance") +
    ggplot2::labs(x = NULL, y = NULL) +
    ggplot2::theme_minimal(base_size = 9) +
    # The left margin holds the first pool's name, leaning left at 60 degrees, at the length of
    # the longest.
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 60, hjust = 1),
                   panel.grid = ggplot2::element_blank(),
                   plot.margin = ggplot2::margin(5.5, 5.5, 5.5, max(5.5, 2 * max(nchar(pools)))))
# The figure grows by the longest pool name, at about 0.05 in a character in its axes, so the
# tiles keep their size.
name_room <- 0.05 * max(nchar(pools))
side <- min(7, 0.25 * length(pools) + 2.5)
report_plot(figure, "distances", width = side + 1.2 + name_room, height = side + 0.87 * name_room,
            caption = "The distance matrix above, from distance.tsv. Blue below zero, red above.")
```

## What each distance rests on

```{r mds-sites, results = "asis"}
few <- num(pairs$few_sites) == 1
if (min(shared) == max(shared)) {
    if (all(few)) {
        cat("**Every pair is averaged over ", report_number(min(shared)), " sites, fewer than ",
            "30** (few_sites is 1 throughout distance.tsv), so every distance here is rough ",
            "rather than wrong.\n\n", sep = "")
    } else {
        cat("Every pair is averaged over ", report_number(min(shared)), " sites (distance.tsv).\n\n",
            sep = "")
    }
} else {
    matrix_tables(pool_matrix("sites"), "Shared sites",
                  paste0("How many sites each pair's distance is averaged over (distance.tsv). A ",
                         "pair rests on the sites both of its pools were read at, so two pairs in ",
                         "one run can rest on different counts."))
}

if (min(shared) != max(shared) && all(few)) {
    cat("**Every pair is averaged over fewer than 30 sites** (few_sites is 1 throughout ",
        "distance.tsv), so every distance here is rough rather than wrong.\n\n", sep = "")
} else if (!all(few) && any(few)) {
    table <- data.frame("Pool A" = pairs$pool_a[few], "Pool B" = pairs$pool_b[few],
                        Sites = report_number(pairs$sites[few]),
                        Distance = report_number(pairs$distance[few], digits = 3),
                        check.names = FALSE, stringsAsFactors = FALSE)
    report_table(table, max_rows = 40, source = "distance.tsv",
                 caption = paste0("The pairs averaged over fewer than 30 sites (few_sites in ",
                                  "distance.tsv). Their distances are rough rather than wrong."))
}

before <- data.frame(raw = num(pairs$raw), corrected = num(pairs$distance),
                     shared = ifelse(few, "fewer than 30 sites", "30 sites or more"))
# Both axes on one scale from zero, so the line of no correction is on the panel however much the
# correction removed.
limits <- range(c(0, before$raw, before$corrected), na.rm = TRUE)
figure <- ggplot2::ggplot(before, ggplot2::aes(x = raw, y = corrected, shape = shared)) +
    ggplot2::geom_abline(slope = 1, intercept = 0, color = "gray60", linetype = "dashed") +
    ggplot2::geom_hline(yintercept = 0, color = "gray80") +
    ggplot2::geom_point(size = 2) +
    ggplot2::scale_shape_manual(values = c("30 sites or more" = 16, "fewer than 30 sites" = 1)) +
    ggplot2::coord_fixed(xlim = limits, ylim = limits) +
    ggplot2::labs(x = "Uncorrected distance", y = "Corrected distance", shape = NULL) +
    ggplot2::theme_bw(base_size = 9) +
    # The right margin takes half of a tick label at the panel's edge.
    ggplot2::theme(legend.position = "bottom", plot.margin = ggplot2::margin(5.5, 16, 5.5, 5.5))
report_plot(figure, "correction", width = 4.6, height = 5,
            caption = paste0("Every pair before and after the sampling correction, from the raw and ",
                             "distance columns of distance.tsv. The dashed line is no correction; ",
                             "how far a point sits below it is what the correction removed."))
```
